\documentclass{article}
\usepackage{tikz-cd}
\begin{document}
\subsection{The idea of a Category}
A category is a mathematical structure that links a collection of objects with non-symmetric relationships called morphisms.
They are used to create abstract models of mathematical concepts and structures.
The objects and morphisms are extremely versatile in what they can represent as shown in the commutative diagrams below:
\begin{equation}
    \begin{tikzcd} 
        A \ar[r,"Father"]
            & B \ar[d,"Brother"] \\
            & C
    \end{tikzcd}
\end{equation}
It is useful to see categories as just items with arrows between them like
above but it should be noted that for each object, there is an identity
morphism that points to itself.  Every time morphisms can be combined, such
as how a parent's brother is an uncle, it is said a composite morphism
exists, a direct connection from you to your uncle.  Compositions and
identities are often not shown as they are redundant information given that
it is known the diagram is meant to be a category.

Abstracting with categories and comparing them can allow us to spot
similarities between situations that intuitively seem very different as
each situation has been abstracted into the same format.  This comparison
may help us understand one of the situations better or try an approach to
proofs or computations around them based on the similarities to the other
category.  For different contexts equivalent concepts could be more easily
translated over, such as how we can think of adding numbers and then
similarly add sets and vectors.  Putting an idea into a category in itself
can also lead to new lines of thought as it is taking away all of the
context and presenting it in a new way.  Abstracting to motivate reasoning
is an essential part of mathematics for instance how algebra has helped us
think about numbers and arithmetic.
\subsection{Axioms and notation}
\subsection{Size of categories}
\subsection{Isomorphisms}
\end{document}